---
title: "Unit 5 Project - Hospital"
author: 
  - "Miguel Zavala - Student"
  - "Ahmand Salam - Student"
  - "Arman Azhand - Student"
date: "2025-01-29"
output: 
  # word_document: default
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
library(ggplot2)
library(visdat)
library(tidyverse)
library(corrplot)
library(GGally)
library(car)
library(ggcorrplot)
library(glmnet)
library(caret)

knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
#rmarkdown::render('HospitalStay.Rmd', output_file = 'index.html')
```

# Hospital Stay

## EDA

### Import Data

```{r}
hospitalData <- read.csv("./project_hospitals.csv", header = TRUE)
```

### Column Descriptions

-   ID: Unique identifier.
-   Lgth.of.Sty: Length of stay (Variable of interest).
-   Age: Average Age of the patient.
-   Inf.Risk: Infection risk.
    -   Average estimated probability of hospital infection
-   R.Cul.Rat: Culture rate.
    -   Ratio of \# of cultures performed to number of patients without symptoms of infection times 100
-   R.CX.ray.Rat: Chest X-ray rate.
    -   Ratio of \# of cultures performed to number of patients without symptoms of pneumonia time 100
-   N.Beds: Number of beds in the hospital.
    -   Average number of beds.
-   Med.Sc.Aff: Medical school affiliation.
    -   1=Yes, 2=No
-   Region: Region of the hospital.
    -   Geographic region: 1=NE, 2=NC,3=S, 4=W
-   Avg.Pat: Average number of patients.
    -   Average number of patients in hospital per day.
-   Avg.Nur: Average number of nurses.
    -   Average number of full time nurses.
-   Pct.Ser.Fac: Percentage of service facilities.
    -   Percent of 35 potential facilities and services that are provided by the hospital

### Sanity check

#### Summary of the Data Set

```{r}
summary(hospitalData)
```

#### Finding empyt values

```{r}
colSums(is.na(hospitalData))
```

##### Plot of missing data

```{r}
vis_miss(hospitalData)
```

Our data set `hospitalData` from the csv `HospitalDurations` contains 113 observations of 12 variables. We have no missing values for any variable, thus we do not need to impute for any observations. Out of the 12 variables, one of them is an identifier variable `ID`, and will not be used for predicting. The patient's length of stay `Lgth.of.Sty` is the variable we would like to predict for. Therefor, we have 10 potential explanatory variables and 1 variable we are trying to predict for.

### Visualizing Hospital Stay Distribution

```{r}
hist(hospitalData$Lgth.of.Sty, xlab = "Lenght of Stay (Days)", main = "")
```

We see some slight right skew for Lgt.of.Sty

### Data Cleaning

```{r}
cleanData <- hospitalData |> select(-ID)
# cleanData$RegionEncoded <- factor(cleanData$Region,
#   levels = c(1, 2, 3, 4),
#   labels = c("NE", "NC", "S", "W")
# )
cleanDataLabels <- c(
  "Length of Stay",
  "Age",
  "Infection Risk",
  "Routine culturing ratio",
  "Routine chest X-ray ratio",
  "Number of beds",
  "Medical School Affiliation",
  "Region",
  "Average Daily census",
  "Number of nurses",
  "Available facilities"
)
```

### Initial Graphs


### Correlation 
```{r}
cor_matrix = cor(cleanData |> select(-Region))

# corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45)
ggcorrplot(cor_matrix, lab = TRUE)
```

Length of Stay correlated more strongly with: + Infection Risk, + Average Patients, - Region (however, this is not a numerical variable but rather a categorical variable and should be discounted)

```{r}
pairs(cleanData |> select(-Region, -Med.Sc.Aff), panel = function(x, y) {
  points(x, y, pch = 16, col = "blue")  # Scatter points
  abline(lm(y ~ x), col = "red")  # Add trend line
})
```

```{r}
ggpairs(cleanData |> select(-Region, -Med.Sc.Aff), lower = list(continuous = wrap("smooth", method = "lm", color = "blue")), progress = FALSE)
```

The variables Number of Beds, Average Patients, Average Nurses, and Percentage of service facilities; exhibit high intercorrelation, particularly between Number of Beds and Average Patients (0.981), which could introduce collinearity in a predictive model.

An important strong correlation that appears to be an important factor in hospital stays is Infection Risk (0.533).

Both Culture Rate and Chest X-Ray Rate displays moderate correlation with Infection Risk; thhis suggests an intuitive relationship: as infection risk increases, more diagnostic tests are performed. Infection Risk may also serve as a useful predictor for Length of Stay, as higher infection risks could lead to prolonged hospital stays.

### Correlograms with Log Transformations
```{r}
corData <- cleanData %>% select(-Med.Sc.Aff, -Region)
lin_logCorData <- log(corData)
lin_logCorData$Lgth.of.Sty <- corData$Lgth.of.Sty
log_linCorData <- corData %>% select(-Lgth.of.Sty)
log_linCorData$logLgth.of.Sty <- log(corData$Lgth.of.Sty)
```

#### Linear-Linear Corr
```{r}
correlation_matrix <- cor(corData, method = "pearson")
ggcorrplot(correlation_matrix, lab = TRUE, title = "Linear-Linear")
```

#### Linear-Log Corr
```{r}
lin_log_correlation_matrix <- cor(lin_logCorData, method = "pearson")
ggcorrplot(lin_log_correlation_matrix, lab = TRUE, title = "Linear-Log")
```

#### Log-Linear Corr
```{r}
log_lin_correlation_matrix <- cor(log_linCorData, method = "pearson")
ggcorrplot(log_lin_correlation_matrix, lab = TRUE, title = "Log-Linear")
```

### Data Viz

```{r}
hist(cleanData$Lgth.of.Sty, xlab = "Length of Stay (Days)", main = "Histogram of Patient's Length of Stay")
```

We see some slight right skew for Lenght of Stay that could suggest

#### Lenght of Stay vs. Infection Risk

```{r}
ggplot(cleanData, aes(y = Inf.Risk, x = Lgth.of.Sty)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Length of Stay vs. Infection Risk")
```

We can spot two outliers around 17 and 20 when we analyze the relationship with Infection Risk those values would consider further analysis to understand the behavior.

#### Lenght of Stay vs. Region
```{r}
ggplot(cleanData, aes(y = Lgth.of.Sty, x = Region)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(title = "Length of Stay vs. Region")
```

#### Lenght of Stay vs. Avg. Nurses
```{r}
ggplot(cleanData, aes(y = Lgth.of.Sty, x = Avg.Nur)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "green") +
  labs(title = "Length of Stay vs, Avg Number of Nurses")
```

We see an increase of infections at lower averages in the number of nurses, with a Plateu effect when reaching close to 200 in average and continuing the trend over 400.

#### Lenght of Stay vs. Avg. Patiernts
```{r}
ggplot(cleanData, aes(y = Lgth.of.Sty, x = Avg.Pat)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "purple") +
  labs(title = "Length of Stay vs. Avg Patients")
```

#### Lenght of Stay vs. Avg. Number of Beds
```{r}
ggplot(cleanData, aes(y = Lgth.of.Sty, x = N.Beds)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "brown") +
  labs(title = "Length of Stay vs Avg. Number of Beds")
```


### Length of Stay vs other continuous variables
```{r}
for (i in c(2:6, 9:11)) {
  tempPlot <- ggplot(data = cleanData, aes(x = cleanData[, i], y = Lgth.of.Sty)) +
    geom_point() +
    geom_smooth() +
    ggtitle(paste0(cleanDataLabels[i], " vs Length of Stay")) +
    xlab(cleanDataLabels[i]) +
    ylab("Length of Stay (Days)") +
    theme(legend.position = "none")
  print(tempPlot)
}
```

### Length of Stay vs log of other continuous variables
```{r}
for (i in c(2:6, 9:11)) {
  tempPlot <- ggplot(data = cleanData, aes(x = log(cleanData[, i]), y = Lgth.of.Sty)) +
    geom_point() +
    geom_smooth() +
    ggtitle(paste0("Log ", cleanDataLabels[i], " vs Length of Stay")) +
    xlab(paste0("log ", cleanDataLabels[i])) +
    ylab("Length of Stay (Days)") +
    theme(legend.position = "none")
  print(tempPlot)
}
```

### log Length of Stay vs other continuous variables
```{r}
for (i in c(2:6,9:11)) {
  tempPlot = ggplot(data=cleanData,aes(x=log(cleanData[,i]),y=log(Lgth.of.Sty))) +
    geom_point() +
    geom_smooth() +
    ggtitle(paste0("Log ", cleanDataLabels[i]," vs Length of Stay")) +
    xlab(cleanDataLabels[i]) +
    ylab("log Length of Stay (Days)") +
    theme(legend.position = "none")
  print(tempPlot)
}
```

## Objective 1

```{r}
allVarsModel = lm(Lgth.of.Sty ~ ., data = cleanData)
par(mfrow = c(2,2))
plot(allVarsModel)
```

### All Variables Model Sumary
```{r}
summary(allVarsModel)
```

The diagnostic plots suggest the linear model fits reasonably well. 

The Residuals vs. Fitted and Scale-Location plots show residuals scattered around zero, with no strong signs of nonlinearity or heteroskedasticity. However, Observation 47 stands out, showing a large residual across multiple plots. In the Q–Q plot, most points follow the regression line, but Observation 47 appears on the right tail, deviating from normality.

Leverage diagnostics indicate that most data points are well-standardized around the fitted line. Observation 47 is near the threshold for Cook’s distance, suggesting potential influence, while Observation 112 exhibits high leverage without strong influence. These points may warrant further investigation to understand their impact on the model.

No major assumption violations are evident, but examining these observations could help assess model robustness.


### Variable Inflation

```{r}
vif(allVarsModel)
```

Number of Beds and Average Patients have *extremely* high VIFs, suggesting high multicollinearity. This makes sense, as the more room a hospital has (number of beds) then the more patients they can have on average. Average number of full time nurses has potential collinearity as well, which also makes sense as a hospital with more beds will likely have more nursing staff to take care of the patients.


```{r Length of Stay vs Infection Risk by Region}

## Interaction Exploration
cleanData$RegionAsFactor <- factor(cleanData$Region, levels = c(1, 2, 3, 4), labels = c("NE", "NC", "S", "W"))
ggplot(cleanData, aes(x = Inf.Risk, y = Lgth.of.Sty, color = RegionAsFactor)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Length of Stay vs Infection Risk by Region") +
  scale_color_manual(values = c("red", "blue", "green", "purple")) +
  labs(color = "Region")
```

### Length of Stay vs Infection Risk by Region Interaction Observations:

NE (Red) shows the steepest slope: Infection risk has a stronger impact on Length of Stay in Northeast hospitals compared to other regions.However, the NE region has more high outliers. 
NC (Blue) and S (Green) have moderate slopes, suggesting that infection risk moderately influences hospital stay duration in these regions.
W (Purple) appears almost flat, low variance. This may indicate that in Western hospitals, infection risk has little to no impact on length of stay.

### Anova Model - Fit ANOVA model with interaction effect
```{r}
anova_model <- lm(Lgth.of.Sty ~ Inf.Risk * RegionAsFactor, data = cleanData)
summary(anova_model)
```

#### ANOVA test for significance of interaction terms
```{r}
anova(anova_model)
```

#### # Checking Variance Inflation Factor (VIF) for multicollinearity
```{r}
vif(anova_model)
```

Interpretations: 

- A one-unit increase in infection risk leads to an increase of ~1.35 days in hospital stay (in the baseline NE region). This is highly statistically significant, suggesting infection risk is a key predictor of length of stay.

Patients in North Central hospitals stay approximately 3.02 days longer than those in the Northeast on average, with a statistically significant result. Similarly, patients in Southern hospitals have an average stay that is about 2.59 days longer than in the Northeast, also showing statistical significance. In the Western region, the average length of stay is around 3.5 days longer than in the Northeast; however, this result is only marginally significant.

```{r}

# Prepare data: Separate predictors (X) and target variable (y)
X <- as.matrix(cleanData[, -which(names(cleanData) == "Lgth.of.Sty")])  # Convert predictors to matrix
y <- cleanData$Lgth.of.Sty  # Target variable

# Perform LASSO regression with cross-validation
cv_lasso <- cv.glmnet(X, y, alpha = 1, nfolds = 10)  # alpha = 1 for LASSO

# Plot cross-validation results
plot(cv_lasso)

# Find the optimal lambda (penalty parameter)
best_lambda <- cv_lasso$lambda.min
print(paste("Optimal lambda:", best_lambda))

# Fit final LASSO model using optimal lambda
lasso_model <- glmnet(X, y, alpha = 1, lambda = best_lambda)

# Extract feature coefficients
lasso_coefficients <- coef(lasso_model)

# Convert coefficients to a named vector
lasso_coefficients <- as.vector(lasso_coefficients)
names(lasso_coefficients) <- rownames(coef(lasso_model))

# Print Intercept
intercept <- lasso_coefficients["(Intercept)"]
print(paste("Intercept:", intercept))

# Print selected features and their coefficients
selected_features <- lasso_coefficients[lasso_coefficients != 0]  # Non-zero coefficients
print("Selected Features and Coefficients:")
print(selected_features)

# Compute RMSE using cross-validation
predictions <- predict(lasso_model, X)  # Predict on training data
rmse <- sqrt(mean((y - predictions)^2))  # RMSE calculation
print(paste("RMSE:", rmse))

# Compute R-squared
ss_total <- sum((y - mean(y))^2)  # Total sum of squares
ss_residual <- sum((y - predictions)^2)  # Residual sum of squares
r_squared <- 1 - (ss_residual / ss_total)  # R-squared

# Compute Adjusted R-squared
n <- length(y)  # Number of observations
p <- length(selected_features) - 1  # Number of selected predictors (excluding intercept)
adj_r_squared <- 1 - ((1 - r_squared) * (n - 1) / (n - p - 1))

print(paste("Adjusted R-squared:", adj_r_squared))

```